---
title: PhD Toolbox 2017 Project
output: html_document
---

### **A reproducible research step by step!** 

#### _Index_
1. _Install the packages we are going to need_
1. _Download and save data_
1. _Data cleaning and processing_
    1. _Remove missing values_
    1. _Extract and keep the most common species, calculate new variables_
1. _Export and save cleaned data_
1. _Plots_
    1. _Scatter and Barplot_
    1. _Time series plot_

##### **1. Install the packages we are going to need** 
```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
# install.packages("tidyverse")   
# install.packages("DT")
library(tidyverse)
library(DT) 
```

##### **2. Download and save data**
```{r surveys}
download.file("https://ndownloader.figshare.com/files/2292169","portal_data_joined.csv")
surveys <-read.csv("portal_data_joined.csv") # Choose your own directory!
```


##### **3. Data cleaning and processing**
###### 3.1. Remove missing values
```{r surveys_complete}
surveys_complete <- surveys %>%
filter(species_id != "",            # remove missing species_id
         !is.na(weight),            # remove missing weight
         !is.na(hindfoot_length),   # remove missing hindfoot_length
         sex != "")                 # remove missing sex
```

###### 3.2. Extract and keep  most common species, calculate new variables
```{r species_counts, results='asis'}
# Extract most common species...
species_counts <- surveys_complete %>%
  group_by(species_id) %>% 
  tally %>%
  filter(n >= 50)

#...and keep them!
surveys_complete <- surveys_complete %>%
  filter(species_id %in% species_counts$species_id)

# Calculate mean and standard deviation of weight... 
surveys_complete <- surveys_complete %>%
  group_by(species)  %>%
  mutate(mean_weight = mean(weight), sd_weight = sd(weight))

# ..and hindfoot lenght
surveys_complete <- surveys_complete %>%
  group_by(species) %>%
  mutate(mean_hindfoot_length = mean(hindfoot_length), sd_hindfoot_length = sd(hindfoot_length)) 
  
# Display dataset in a clearer way than that produced using the command *head* 
datatable(surveys_complete[1:1000,1:17]) 
```


##### **4. Export and save cleaned data in a new .csv file**
```{r surveys_cleaned}
write.csv(surveys_complete, file = "surveys_cleaned.csv", row.names=FALSE)
```


##### **5. Plots**
###### **5.1. Scatter and Barplot**
```{r surveys_plot}
surveys_plot <- read.csv ("surveys_cleaned.csv", stringsAsFactors = F)

# Let's set a theme that will be used in all of the plots
mytheme <- theme_bw() +
           theme(panel.grid.major.x = element_blank(), 
           panel.grid.minor.x = element_blank(), 
           panel.grid.major.y = element_blank(), 
           panel.grid.minor.y = element_blank())

# Produce a scatterplot
plot1 <- ggplot(data = surveys_plot, # create background
         aes(x = weight, y = hindfoot_length)) # create x and y axes

# Plot points with different colors and shapes for each species 
plot1 <- plot1 + geom_jitter(aes(color = species)) +
         facet_grid(. ~ sex) + # split the plot by sex
         mytheme +  
         labs(title = 'Observed species measures', x = 'Weight', y = 'Hindfoot lenght') + 
         theme(text = element_text(size=10, family = "Avenir")) # change font and size
plot1 

### Produce a barplot of mean weight per species
dodge <- position_dodge(width = 0.5) 

weight_plot <- ggplot(data = surveys_plot, 
               aes(x = species, y = weight, fill = species))

weight_plot <- weight_plot + geom_bar(stat = "identity", position = dodge, width = 0.5) +
               geom_errorbar(aes(ymin = mean_weight - sd_weight, ymax = mean_weight + sd_weight), 
                             color = "black", position = dodge, width = 0.25) +
               facet_grid(. ~ sex) +
               mytheme + 
               labs(x = 'Species', y = 'Weight') + 
               theme(text = element_text(size = 10, family = "Avenir")) + # change font and size
               theme(axis.text.x = element_text(angle = 60, hjust = 1))     # change labels orientation
weight_plot

### Produce a barplot of mean hindfoot lenght per species
hindfoot_plot <- ggplot(data = surveys_plot, 
                 aes(x = species, y = hindfoot_length, fill = species))

hindfoot_plot <- hindfoot_plot + geom_bar(stat="identity", position=dodge, width=0.5) +
                 geom_errorbar(aes(ymin = mean_hindfoot_length - sd_hindfoot_length, 
                                   ymax = mean_hindfoot_length + sd_hindfoot_length), color = "black", 
                                   position = dodge, width = 0.25) +
                 facet_grid(. ~ sex) +
                 mytheme + 
                 labs(x = 'Species', y = 'Hindfoot lenght') + 
                 theme(text = element_text(size = 10, family ="Avenir")) +
                 theme(axis.text.x = element_text(angle = 60, hjust = 1)) 
hindfoot_plot 
```


###### **5.2. Time series Plot**
```{r yearly_sex_counts}

# To plot an annual time series data we need to group data
# Create a new dataset based on sex, year and species
yearly_sex_counts <- surveys_complete %>% 
  group_by(sex, year, species) %>%
  tally

timeseries_plot <- ggplot(data = yearly_sex_counts, 
                   aes(x = year, y = n, color = species, group = species)) 

timeseries_plot <- timeseries_plot +  geom_line(aes(group = species, color = species)) +
                   facet_grid (. ~ sex) +
                   mytheme +
                   labs(title = 'Observed species in time', x = 'Year of observation', y = 'Number of species') +
                   theme(text = element_text(size = 10, family = "Avenir"))         
timeseries_plot
```
